Numerical modelling of the Cordilleran ice sheet

### 2014-07-18 kappa.tex
Polish physical setting.

During the last glacial cycle, much of North America has been covered by ice
(\cref{fig:map-northamerica}).
During the Last Glacial Maximum (LGM), three major ice sheets had collided to
form a continuous blanket of ice that spanned from southern Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Ohio
(\cref{fig:map-northamerica}). The Laurentide ice
sheet, largest of the three, was centred on the Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, \emph{the Cordilleran ice sheet}, intermediate in size,
capped the mountain
ranges of the Western Cordillera.

@@

The North American Cordillera forms an elongated but broad mountain system with
an average east-west width of about 1000\,km (\cref{fig:map-cordillera}). Its
topography is a complex jigsaw of mountain ranges and interior basins. It was
formed during multiple accretion phases associated with the closure of several
back-arc oceanic trenches, over a 200 million year (Ma)-long geological history
\citep{Sigloch.Mihalynuk.2013}.

The dominant reliefs occur along two major ribbons that run
parallel to each other along the general orientation of the North American Cordillera
(\cref{photo-alaska-range}). The western ribbon is formed by the Pacific Coast
orogenic belt \mref\ and includes,
from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench \mref, depression and includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky mountains in the South. The
Brooks Range of northern Alaska has been covered by ice, but its glaciers
remained disjunct from the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

@@

The North American Cordillera hosts several intermontane basins, the presence
of which influenced former patterns of Cordilleran ice sheet flow. The two
major depressions are the Fraser Plateau in south-central British Columbia, and
the Liard Lowland in northern British Columbia (\cref{fig:map-cordillera}).
These two basins are separated by the Skeena Mountains, a mildly elevated, but
extensive and deeply incised mountain group. This is the most dominant upland
area in-between the dominant topographic ribbons that form the North American
Cordillera.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effects on
vegetation are clearly expressed in the landscape
(\cref{fig:photo-fraser-valley}). Due to the large latitudinal extent of the
region, \emph{mean annual temperatures} vary significantly from north to south
(\cref{fig:plot-atm}a). About half of the model domain has mean annual
temperatures below freezing. However, \emph{temperature seasonality}\renote{
    This is a common term in the meteorological literature, I think. At least
    this is what Qiong told me.},
here computed as the difference between the warmest and the coldest months,
varies between around 10\,\degC\ along most of the Pacific Coast to above
40\,\degC\ for the interior Arctic sectors \renote{
    I think the word ``polar'' or ``Arctic'' is needed here, see
    \cref{fig:plot-atm}b.}
(\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperatures are well below
freezing point experience warm summers.

\emph{Annual precipitation rates} are highest on the western slopes of the
Pacific Coast ranges and decline quickly on the eastern side of the mountains
(\cref{fig:photo-fraser-valley}, \cref{fig:plot-atm}c). In terms of ice sheet
mass balance, this contrast is even reinforced through the difference in timing
of the \emph{precipitation peaks} through the year (\cref{fig:plot-atm}d).
Coastal regions receive most precipitation during winter and fall, which is
therefore largely delivered as snow. In contrast, many inland regions
experience dry winters, and most of the annual precipitation, consequently
falls there as rain during summer, further enhancing ice melt. The climatic
setting of the North American Cordillera is further described in \CCLI, where
its effects on the ice sheet model response have been studied in detail.

@@

Evidence from glaciofluvial deposited dated using cosmogenic nuclides at $2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the Cordilleran mountains in the early Pleistocene
\citep{Hidy.etal.2013}.

@@

In the \emph{northern sector} of the Cordilleran ice sheet domain, at least four glaciations have
been identified
    \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
           Briner.Kaufman.2008, Demuro.etal.2012,
           Stroeven.etal.2010, Stroeven.etal.2014}.
The two older ones, often collectively referred to as the Reid  and pre-Reid glaciations,
pre-date the last glacial cycle, and are therefore beyond the scope of this study.
The two younger glaciations, however, occurred both during the last glacial
cycle. First, the older of the two, the Gladstone Glaciation, reached its maximum extent between
$50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
during Marine Oxygen Isotope Stage (MIS) 4, a period of relatively large global ice volume. However, evidence from the Gladstone Glaciation has not been
documented along the southern ice sheet margin, making it difficult to
seize its spacial extent. The younger of the two last glacial cycle glaciations, the McConnell Glaciation, is associated with end
moraines of oldest minimum apparent exposure ages of
$15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
which appears to correspond to the last global ice volume maximum (MIS~2). In contrast to
the Gladstone Glaciation, the McConnell Glaciation appears contemporaneous with glacial advance of the southern ice sheet margin during the Fraser
glaciation, with a maximum extent attained between
$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep{Porter.Swanson.1998}.
Within errors inherent to the cosmogenic nuclide exposure dating methods \citep{Heyman.etal.2011}, the
McConnell and Fraser glaciations correspond to the same glacial episode, which
resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet. Apart from geological evidence of the final demise of the ice sheet, most field evidence relate to the LGM maximum extent.

Ice has covered most of the
Canadian Cordillera, and some of the adjacent regions in north-western contiguous
USA and Alaska (\cref{fig:map-cordillera}). The ice sheet originated from the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}. It is generally thought that the LGM configuration of the Cordilleran ice sheet was short-lived
    \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a hypothesis which can be studied through numerical modelling.

The \emph{southern margin} of the ice sheet has been studied extensively, and
is in several respects better understood than its northern counterpart
\citep{Booth.etal.2003}. It consists of coalescent piedmont
lobes, each of which formed at the outlet of the main valleys. The Puget Lobe
\citep{Thorson.1980, Porter.Swanson.1998} was the westernmost and
largest Lobes. It covered the Puget Lowland, a region that comprises
the cities of Vancouver and Seattle. Further east, beyond the Cascade Range, the
Okanagan and Purcell lobes formed the southern Cordilleran ice sheet margin. The latter once dammed the Clark Fork River, inhibiting runoff and forming
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet margin retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape, known as the Channeled Scabland.

The \emph{western margin} of the Cordilleran ice sheet covered part of the Pacific continental
shelf. An independent
ice cap covering Vancouver Island diverted flow from the main body of ice to drain into the Strait of Georgia between Vancouver Island and the Olympic Peninsula to form the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
Further to the North, the ice margin encroached the Queen Charlotte Islands. However, the
archipelago may not have been fully ice-covered, even during maximum of glaciation. Submarine troughs traversing several portions of the continental shelf indicate the erosional action of
additional ice streams, yet it is not clear whether they have formed during the
last glacial cycle or during earlier glaciations \mref.

Finally, the conditions at the \emph{eastern margin} of the Cordilleran ice sheet are the least understood.
Although it
is known that the Cordilleran ice sheet collided with the Laurentide ice sheet at times during the last glacial cycle,
the nature of the interaction between the two ice sheets and the precise
location of this junction remains unclear \citep{Gowan.2013}, while limited information exists on its
timing
    \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}.
Because the
present thesis focuses specifically on the Cordilleran ice sheet, I, therefore, neglect buttressing effect of the
Laurentide ice sheet.

The retreat of the ice sheet during the last deglaciation was likely rapid. Most of the evidence for the style and pace of deglaciation comes
from the Fraser Plateau. Studies here have indicated that some of the highest relief
emerged from the ice prior to the surrounding lowlands, possibly resulting in thin,
stagnant ice remnants covering parts of the plateau
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
Less is known about the deglaciation of the northern sector of the Cordilleran ice sheet, but abundant glacial landforms from the Liard Lowland indicate a
different mode of \emph{deglaciation}, with active retreat of the ice margin
towards the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in \CCYC.

\todo{To make this more complete I suggest two additions: 1. a few sentences regarding the timing of deglaciation (to complement your introduction where you have timing of glaciation) - look in my 2010 and Martin's papers for information. 2. A few more references older than Margold et al. 2013. You might check what Ward has written fordeglaciation in the Yukon for example? Didn't Dyke write about the deglaciation of the Liard Lowland?}

>>>>

During the last glacial cycle, much of North America has been covered by ice
(\cref{fig:map-northamerica}). During the Last Glacial Maximum (LGM), peak of
glaciation, three major ice sheets had collided to
form a continuous blanket of ice that spanned from southern Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago\renote{
    Although already a few 100\,km into the ice margin, I think that Chicago
    is a more illustrative location than Ohio or smaller cities there.}.
The Laurentide ice sheet, largest of the three, was centred on the Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the \emph{Cordilleran ice sheet}, intermediate in size,
capped the mountain
ranges of the Western Cordillera.

@@

The North American Cordillera forms an elongated but broad mountain system with
an average east-west width of about 1000\,km (\cref{fig:map-cordillera}). Its
topography is a complex jigsaw of mountain ranges and interior basins. It was
formed during multiple accretion phases associated with the closure of several
back-arc oceanic trenches, over a 200 million year (200\,Ma)-long geological
history \citep{Sigloch.Mihalynuk.2013}.

The \emph{dominant reliefs} occur along two major ribbons that run parallel to
each other along the general orientation of the North American Cordillera. The
western ribbon is formed by the Pacific Coast orogenic belt and includes, from
North to South, the Alaska Range (\cref{fig:photo-alaska-range}), the Wrangell
and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench depression, and includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky mountains in the South. The
Brooks Range of northern Alaska has been covered by ice, but its glaciers
remained disjunct from the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

@@

The North American Cordillera hosts several \emph{intermontane basins}, the
presence
of which influenced former patterns of Cordilleran ice sheet flow. The two
major depressions are the Fraser Plateau in south-central British Columbia, and
the Liard Lowland in northern British Columbia (\cref{fig:map-cordillera}).
These two basins are separated by the Skeena Mountains, a mildly elevated, but
extensive and deeply incised mountain group. This is the most dominant upland
area in-between the dominant topographic ribbons that form the North American
Cordillera.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effects on
vegetation are clearly expressed in the landscape
(\cref{fig:photo-fraser-valley}). Due to the large latitudinal extent of the
region, mean annual \emph{temperatures} vary significantly from north to south
(\cref{fig:plot-atm}a). About half of the model domain has mean annual
temperatures below freezing. However, temperature \emph{seasonality}\renote{
    ``Seasonality'' is a common term in the meteorological literature,
    implicitly referring to temperature. At least this is what I understood
    from Qiong.},
here computed as the difference between the warmest and the coldest months,
varies between around 10\,\degC\ along most of the Pacific Coast to above
40\,\degC\ for the interior Arctic sectors\renote{
    I think the word ``polar'' or ``Arctic'' is needed here, see
    \cref{fig:plot-atm}b.}
(\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperatures are well below
freezing point experience warm summers.

Annual \emph{precipitation rates} are highest on the western slopes of the
Pacific Coast ranges and decline quickly on the eastern side of the mountains
(\cref{fig:photo-fraser-valley,fig:plot-atm}c). In terms of ice sheet
mass balance, this contrast is even reinforced through the differences in
\emph{timing of maximum precipitation} through the year (\cref{fig:plot-atm}d).
Coastal regions receive most precipitation during winter and fall, which is
therefore largely delivered as snow. In contrast, many inland regions
experience dry winters, and most of the annual precipitation, consequently
falls there as rain during summer, further enhancing ice melt. The climatic
setting of the North American Cordillera is further described in \CCLI, where
its effects on the ice sheet model response have been studied in detail.

@@

Evidence from glaciofluvial deposits dated using cosmogenic nuclides at
$2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the
Cordilleran mountains during the early Pleistocene \citep{Hidy.etal.2013}.

@@

In the \emph{northern sector} of the Cordilleran ice sheet domain, at least
four glaciations have been identified
    \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
           Briner.Kaufman.2008, Demuro.etal.2012,
           Stroeven.etal.2010, Stroeven.etal.2014}.
The two older ones, often collectively referred to as the Reid and pre-Reid
glaciations, pre-date the last glacial cycle, and are therefore beyond the
scope of this study. The two younger glaciations, however, occurred both during
the last glacial cycle. First, the older of the two, the Gladstone Glaciation,
reached its maximum extent between $50.4\pm3.0$ and $54.3\pm2.0\,^{10}$Be\,ka
\citep{Ward.etal.2007}, during Marine Oxygen Isotope Stage (MIS) 4, a period of
relatively large global ice volume. However, evidence from the Gladstone
Glaciation has not been documented along the southern ice sheet margin, making
it difficult to seize its spacial extent. The younger of the two last glacial
cycle glaciations, the McConnell Glaciation, is associated with end moraines of
oldest minimum apparent exposure ages of $15.7\pm1.5$ and
$17.7\pm1.6\,^{10}$Be\,ka \citep{Stroeven.etal.2014}, which appears to
correspond to the last global ice volume maximum (MIS~2). In contrast to the
Gladstone Glaciation, the McConnell Glaciation appears contemporaneous with
glacial advance of the southern ice sheet margin during the Fraser glaciation,
with a maximum extent attained between $17.4$ and
$16.4$\,\unit{^{14}C\,cal\,ka} \citep{Porter.Swanson.1998}. Within errors
inherent to the cosmogenic nuclide exposure dating methods
\citep{Heyman.etal.2011}, the McConnell and Fraser glaciations correspond to
the same glacial episode, which resulted in the LGM extent of the Cordilleran
ice sheet.

Apart from geological evidence of the final demise of the ice sheet, most field
evidence relate to the LGM extent. It indicates that ice covered most of the
Canadian Cordillera, and some of the adjacent regions in north-western
contiguous USA and Alaska (\cref{fig:map-cordillera}). The Cordilleran ice
sheet originated from the coalescence of several mountain-centred ice caps
\citep{Davis.Mathews.1944}. It is generally thought that this LGM configuration
was short-lived \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a
hypothesis which can be studied through numerical modelling.

The \emph{southern margin} of the ice sheet has been studied extensively, and
is in several respects better understood than its northern counterpart
\citep{Booth.etal.2003}. It consists of coalescent piedmont lobes, each of
which formed at the outlet of the main valleys. The Puget Lobe
\citep{Thorson.1980, Porter.Swanson.1998} was the westernmost and largest one.
It covered the Puget Lowland, a region that comprises the cities of Vancouver
and Seattle. Further east, beyond the Cascade Range, the Okanagan and Purcell
lobes formed the southern Cordilleran ice sheet margin. The latter once dammed
the Clark Fork River, inhibiting runoff and forming Glacial Lake Missoula
\citep{Pardee.1910}. As the ice sheet margin retreated, the lake repeatedly and
catastrophically drained to the Pacific Ocean \citep{Bretz.1923,Waitt.1980}.
These floods contributed to shape a distinctive landscape, known as the
Channeled Scabland.

The \emph{western margin} of the Cordilleran ice sheet covered part of the
Pacific continental shelf. An independent ice cap covering Vancouver Island
diverted flow from the main body of ice to drain into the Strait of Georgia
between Vancouver Island and the Olympic Peninsula, where it formed the Juan
the Fuca marine ice lobe, calving into the Pacific Ocean \mref. Further to the
North, the ice margin encroached the Queen Charlotte Islands. However, the
archipelago may not have been fully ice-covered, even during the maximum of
glaciation. Submarine troughs traversing several portions of the continental
shelf indicate the erosional action of additional ice streams, yet it is not
clear whether they have formed during the last glacial cycle or during earlier
glaciations \mref.

\todo{Find missing refs}

Finally, the conditions at the \emph{eastern margin} of the Cordilleran ice
sheet are the least understood. Although it is known that the Cordilleran ice
sheet collided with the Laurentide ice sheet at times during the last glacial
cycle \citep[e.g.][]{Margold.etal.2013, Margold.etal.2013a}, the nature of the
interaction between the two ice sheets and the precise location of this
junction remains unclear \citep{Gowan.2013}, while limited information exists
on its timing \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}. Because
the present thesis focuses specifically on the Cordilleran ice sheet, I,
therefore, neglect buttressing effects of the Laurentide ice sheet.

The retreat of the ice sheet during the last \emph{deglaciation} was likely
rapid. Most of the evidence for the style and pace of deglaciation comes from
the Fraser Plateau. Studies there have indicated that some of the highest
reliefs emerged from the ice sheet prior to the surrounding lowlands, possibly
resulting in thin, stagnant ice remnants covering parts of the plateau
\citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}. Less
is known about the deglaciation of the northern sector of the Cordilleran ice
sheet, but abundant glacial landforms from the Liard Lowland indicate a
different mode of deglaciation, with active retreat of the ice margin towards
the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in
\CCYC\renote{
    I have not had time to dig into new literature. The Ward paper on
    deglacial valley fills in the Pelly River is like Chinese to me, and I
    could not access the Dyke paper on the Liard Lowland, assuming this is the
    one cited by Martin in his paper.}.
